Gap formation and stability in non-isothermal protoplanetary discs

### 2015-03-27 paper/model.tex

The disc evolution is governed by the standard fluid equations  
\begin{align}\label{3d_gov_eq}
  &\frac{\p\Sigma}{\p t}+\nabla\cdot(\Sigma \bv)=0, \\
  & \frac{\p\bv}{\p t}+\bv\cdot\nabla\bv= -\frac{1}{\rho}\nabla p 
  - \nabla{\Phi} + \bm{f}_\nu,\\
  & \frac{\p e}{\p t} + \nabla\cdot(e\bv) = -p\nabla\cdot\bv +
  \mathcal{H} - \mathcal{C}, %\\
%  &\nabla^2\Phi_d = 4\pi G \Sigma \delta(z),\label{poisson}
\end{align}
where $\Sigma$ is the surface density, $\bv = (v_r,v_\phi)$ the fluid
velocity, $p$ is the vertically-integrated pressure, $e=p/(\gamma-1)$ is the energy
per unit area and the adiabatic index $\gamma=1.4$ is assumed
constant.

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) 
and indirect potentials to account for the non-inertial reference
frame. 
In the momentum equations, $\bm{f}_\nu$ represent viscous forces, 
which includes artificial bulk viscosity to handle shocks, and a
Navier-Stokes viscosity whose magnitude is  
characterised by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$.

\subsection{Heating and cooling}
In the energy equation, the heating term $\mathcal{H}$ is defined as 
\begin{align}
  \mathcal{H} \equiv Q^+ - Q^+_i\frac{\Sigma}{\Sigma_i}, 
\end{align}
where $Q^+$ represents viscous heating (from both physical and
  artificial viscosity) and subscript $i$ denotes
evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as
\begin{align}
  \mathcal{C} \equiv \frac{1}{t_c}\left(e -
  e_i\frac{\Sigma}{\Sigma_i}\right),  
\end{align}
where $t_c = \beta\Omega_k^{-1}$ is the cooling time,
$\Omega_k=\sqrt{GM/r^3}$ is the Keplerian frequency and $\beta$ is an
input parameter. This cooling prescription allows one 
to explore the full range of thermodynamic response of the disc in a 
systematic way: $\beta\ll1$ is a locally isothermal disc while
$\beta\gg1$ is an adiabatic disc.

@@

\subsubsection{Rossby number}
The Rossby number
\begin{align}
  Ro = \frac{{\bf \hat{z}} \cdot \nabla \times \bv - \langle
    {\bf\hat{z}} \cdot \nabla \times \bv
    \rangle_{\phi}}{2\langle\Omega\rangle_{\phi}},  
\end{align}
is a dimensionless measure of relative vorticity. 
 Here $\langle \cdot \rangle_{\phi}$ denotes an azimuthal
average. Values of $Ro<0$ correspond to anti-cyclonic rotation with
respect to the background shear and thus can be used to identify
vortices and quantify its intensity.

>>>>

The disc evolution is governed by the standard fluid equations  
\begin{align}\label{3d_gov_eq}
  &\frac{\p\Sigma}{\p t}+\nabla\cdot(\Sigma \bv)=0, \\
  & \frac{\p\bv}{\p t}+\bv\cdot\nabla\bv= -\frac{1}{\rho}\nabla p 
  - \nabla{\Phi} + \bm{f}_\nu,\\
  & \frac{\p e}{\p t} + \nabla\cdot(e\bv) = -p\nabla\cdot\bv +
  \mathcal{H} - \mathcal{C}, %\\
%  &\nabla^2\Phi_d = 4\pi G \Sigma \delta(z),\label{poisson}
\end{align}
where $\Sigma$ is the surface density, $\bv = (v_r,v_\phi)$ the fluid
velocity, $p$ is the vertically-integrated pressure, $e=p/(\gamma-1)$ is the energy
per unit area and the adiabatic index $\gamma$ is assumed
constant. We take $\gamma=1.4$ unless otherwise stated.

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) 
and indirect potentials to account for the non-inertial reference
frame. 
In the momentum equations, $\bm{f}_\nu$ represent viscous forces, {\bf
which includes artificial bulk viscosity to handle shocks, and a
Navier-Stokes viscosity} whose magnitude is  
characterised by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$.

\subsection{Heating and cooling}
In the energy equation, the heating term $\mathcal{H}$ is defined as 
\begin{align}
  \mathcal{H} \equiv Q^+ - Q^+_i\frac{\Sigma}{\Sigma_i}, 
\end{align}
where $Q^+$ represents viscous heating {\bf(from both physical and
  artificial viscosity)} and subscript $i$ denotes
evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as
\begin{align}
  \mathcal{C} \equiv \frac{1}{t_c}\left(e -
  e_i\frac{\Sigma}{\Sigma_i}\right),  
\end{align}
where $t_c = \beta\Omega_k^{-1}$ is the cooling time,
$\Omega_k=\sqrt{GM/r^3}$ is the Keplerian frequency and $\beta$ is an
input parameter. This cooling prescription allows one 
to explore the full range of thermodynamic response of the disc in a 
systematic way: $\beta\ll1$ is a locally isothermal disc while
$\beta\gg1$ is an adiabatic disc.

@@

\subsubsection{Rossby number}
The Rossby number
\begin{align}
  Ro = \frac{{\bf \hat{z}} \cdot \nabla \times \bv - \langle
    {\bf\hat{z}} \cdot \nabla \times \bv
    \rangle_{\phi}}{2\langle\Omega\rangle_{\phi}},  
\end{align}
is a dimensionless measure of relative vorticity. 
% the ratio of interial to Corriolis forces. 
Here $\langle \cdot \rangle_{\phi}$ denotes an azimuthal
average. Values of $Ro<0$ correspond to anti-cyclonic rotation with
respect to the background shear and thus can be used to identify
vortices and quantify its intensity.
